---
title: 'Homework 1: Categorical Data'
author: "Nicholas G Reich, for PUBHLTH 490ST at UMass-Amherst"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, eval=FALSE)
```

Your assignment must be handed in as a PDF file, reproducibly compiled using LaTeX or markdown.

## Inference for binomial parameter

## Question 1
Say you flip a coin 100 times and get 53 heads. 

a. Write down the Score test statistic and by hand (i.e. don't used a canned R function to calculate the test statistic for you) conduct a test of the null hypothesis $H_0: \pi$=0.5. 
b. Draw a picture of the likelihood and show what features of the likelihood are used for the Score test statistic. 
c. Explain (using pictures and equations)  how the Score test uses different information than the likelihood ratio test.

## Question 2
Conduct a simulation study to evaluate the coverage probabilities of different methods for computing confidence intervals for $\pi$. You may vary the true $\pi$, the sample size, and must choose at least 4 methods to compare. Your results should be summarized with a one paragraph description of the implementation of the simulation study, a one paragraph descirption of the results, and one figure (multi-panel is ok) summarizing the results.

## Question 3
Prove that if $Y\sim Binom(n, \pi)$ and $\pi\sim beta(\alpha_1, \alpha_2)$ (for $\alpha_1>0$ and $\alpha_2>0$) then the posterior $h(\pi|y)$ is a beta distribution, specifically, $h$ follows a $beta(y + \alpha_1 , n - y + \alpha_2)$.

## Question 4
Complete problem 1.17 in CDA.

## Getting started
Let's use the FEV data again and focus on the relationship betwen the fev variable with the age, height, and smoke covariates. To start, let's load the data into our current R session:
```{r, eval=TRUE}
library("Hmisc")
getHdata(FEV)
```

