---
title: "Untitled"
author: "Nicholas G Reich"
date: "September 20, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The dataset we will be using is the titanic3 dataset which is part of the PASWR package. This dataset describes demographic characteristics and outcomes for passengers of the Titanic.


### Loading the data
```{r, message=FALSE}
library(PASWR)
library(arm)
data(titanic3)
str(titanic3)
```


## testing nested models

## looking at residuals

## plotting predicted probabilities for different groups with SEs

## looking at/interpreting coefficients

## running a model with rstanarm

## making likelihood-based CIs?

## looking at leverage


```{r fig.width=10, fig.height=6}
m1 <- glm(survived ~ 1, data=titanic3, family=binomial)


binnedplot(x = predict(m1), y = resid(m1), nclass = 10)

titanic3$preds <- fitted(m1, type="response")
qplot(pclass, preds, data=titanic3, color=pclass, geom=c("jitter")) + 
        ylim(0,1) +
        ylab("predicted probability")
```

The final figure need not have an axis for every variable included in the model, but it should show several different dimensions (using axes, colors, facets, heat maps, etc...).

